---
title: "Assignment 3"
author: "Julie Bang Mikkelsen (AU718507)"
date: "January 7th 2025"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Preprocessing

## Loading required packages

```{r libraries, warning = FALSE, message = FALSE}
library(tidyr)
library(stringr)
library(ggplot2)
library(dplyr)
library(ggspatial)
library(sf)
library(raster)
library(leaflet)
library(rnaturalearth)
library(rnaturalearthdata)
```

## Loading data

I want to investigate whether suicide terrorism committed by female terrorists is more lethal than male terrorists. To do that I use the GTD dataset.

```{r data}
raw_data <- read.csv("globalterrorismdb_0522dist.csv", sep = ";", dec = ",")
#str(raw_data)
```

## Filtering and cleaning data

I filter the data for only suicide attacks, code a gender variable from the summaries, and standardize the outcome variable.

```{r}
female_keywords <- c("female", "woman", "women", "girl", "girls") # preparing list of keywords to identify female terrorists by

d <- raw_data %>% # saving into new df
  filter(suicide == 1) %>% # filtering for suicide attacks
  mutate(female = ifelse(str_detect(tolower(summary), paste(female_keywords, collapse = "|")), 1, 0)) # creating gender variable

d$female <- as.factor(d$female) # converting to factor for better plots
d$region_txt <- as.factor(d$region_txt)

d$nkillwound <- (d$nkill+d$nwound) # 'nkillwound' as total casualties
d$lethality <- d$nkill/(d$nkillwound) # creating outcome variable 'lethality' as ratio killed of total casualities
```

# Analysis

## Visually inspecting overall data

```{r}
# Histogram of lethality
hist(d$lethality)

# Bar plot
ggplot(d, aes(x = iyear, fill = female)) +
  geom_bar() +
  labs(
    x = "Year",
    y = "Number of Attacks",
    fill = "Gender",
    title = "Bar plot of Yearly Attacks") +
  scale_fill_manual(
    values = c("0" = "#e66101", "1" = "#5e3c99"), 
    labels = c("Male/Unclassified", "Female")) +
  theme_minimal()

# Map
d_clean <- d %>%
  filter(!is.na(longitude) & !is.na(latitude)) # Remove rows with NA in coordinates

d_map <- st_as_sf( # Converting clean df to sf object
  d_clean,
  coords = c("longitude", "latitude"),
  crs = 4326) # choosing EPSG:4326 as coordinate system

world <- ne_countries(scale = "medium", returnclass = "sf") # loading world map from naturalearth as sf object

ggplot() +
  geom_sf(data = world, fill = "gray", color = "white") +
  geom_sf(data = d_map, aes(size = nkill, color = female)) +
  labs(
    title = "Geographical Attack Distribution by Gender",
    color = "Gender",
    size = "Casualties") +
  ggspatial::annotation_north_arrow(location = "br") +
  scale_color_manual(
    values = c("0" = "#e66101", "1" = "#5e3c99"),
    labels = c("Male/Unclassified", "Female")) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal")
```

Visually inspecting the data, it appears highly imbalanced. Lethality is not normally distributed, and recent years, male perpetrators, and certain geographical regions are over-represented.

## Visually inspecting for potential predictors of lethality

```{r, fig.height=3.5, fig.width=6}
# Box plot of gender
ggplot(d, aes(x = female, y = lethality, fill = female)) +
  geom_boxplot() +
  labs(
    x = NULL,
    y = "Lethality",
    title = "Boxplot of Lethality by Gender") +
  scale_fill_manual(
    values = c("0" = "#e66101", "1" = "#5e3c99"), 
    labels = c("Male/Unclassified", "Female")) +
  scale_x_discrete(labels = c("0" = "Male/Unclassified", "1" = "Female")) +
  theme_minimal() +
  theme(legend.position = "none")

# Scatter plot of year
ggplot(d, aes(iyear, lethality, col = female)) +
  geom_point(position = position_jitter(width = 0.2, height = 0), alpha = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  labs(
    title = "Scatter plot of Lethality Over Time by Gender",
    x = "Year",
    y = "Lethality",
    color = "Gender") +
  scale_color_manual(
    values = c("0" = "#e66101", "1" = "#5e3c99"),
    labels = c("Male/Unclassified", "Female")) +
    theme_minimal()
```

Inspecting 'gender' alone, there is no visual difference in 'lethality.' However, when also introducing 'year', the scatter plot indicates that there could be a gender effect.

## Correlation testing for year-lethality relationship

```{r}
cor.test(d$iyear, d$lethality, method = "spearman") # Running a correlation on year and lethality
```

Judging from the correlation results, there is a weak but significant (p \< .001) correlation between year and lethality.

## Fitting linear models with year, gender, and geographical region as predictors

```{r}
# transforming year variable to be zero-indexed for interpretability
d <- d %>%
  mutate(iyear_adj = iyear - 1970)

# Fitting models with increasing complexity
m1 <- lm(lethality ~ female, d)

m2 <- lm(lethality ~ female + region_txt, d)
m3 <- lm(lethality ~ female * region_txt, d)

m4 <- lm(lethality ~ iyear_adj + female, d)
m5 <- lm(lethality ~ iyear_adj + female + region_txt, d)
m6 <- lm(lethality ~ iyear_adj + female * region_txt, d)

# Comparing models
anova(m1, m2, m3, m4, m5, m6)
AIC(m1, m2, m3, m4, m5, m6)
```

Comparing the models, the most complex model appears to be the best. m6 is significantly better than the null model, and results in the lowest amount of error and also the lowest AIC-value.

## Inspecting model fit

Baseline indicates male (female = 0) and year = 1970.

```{r}
summary(m1) # inspecting significance of gender alone
summary(m6) # inspecting best model
```

Inspecting the model summary, increasing year and female perpetrators significantly increase attack lethality (p \< .001, and p \< .01, respectively) compared to the baseline.

## Visually presenting interaction effect of geopraphical region

```{r}
# manually creating df with interaction effects from m6
region_interaction <- data.frame(
  region = c("Central Asia", "East Asia", "Eastern Europe", 
             "Middle East & North Africa", "North America", 
             "South America", "South Asia", "Southeast Asia", 
             "Sub-Saharan Africa", "Western Europe"),
  p_value = c(0.27198, NA, 0.75922, 1.07e-06, 0.76285, 
              NA, 0.02596, 0.08764, NA, NA))

# making col with significant/not significant (alfa = .05)
region_interaction$significance <- ifelse(
  region_interaction$p_value < 0.05, "Significant", "Not Significant")
region_interaction$significance[is.na(region_interaction$p_value)] <- "Not Applicable" # primarily regions with too few entries

# preparing translation-table for regions before merging
region_mapping <- c(
  "Central Asia" = "Europe & Central Asia",
  "East Asia" = "East Asia & Pacific",
  "Eastern Europe" = "Europe & Central Asia",
  "Middle East & North Africa" = "Middle East & North Africa",
  "North America" = "North America",
  "South America" = "Latin America & Caribbean",
  "South Asia" = "South Asia",
  "Southeast Asia" = "East Asia & Pacific",
  "Sub-Saharan Africa" = "Sub-Saharan Africa",
  "Western Europe" = "Europe & Central Asia")

# applying translation-table to interaction output df
region_interaction$region_wb <- region_mapping[region_interaction$region]

# merging interaction output df with world df
world_with_significance <- world %>%
  left_join(region_interaction, by = c("region_wb" = "region_wb")) # defining to merge where this variable contains the same value

# replacing NAs with "Not Applicable" in significance col
world_with_significance$significance <- replace_na(world_with_significance$significance, "Not Applicable")

# Map
ggplot() +
  geom_sf(data = world_with_significance, aes(fill = significance), color = "white") +
  labs(
    fill = "Interaction Effect",
    title = "Gender:Region Interaction Map") +
  scale_fill_manual(
    values = c("Significant" = "#fdb863", "Not Significant" = "#8073ac", "Not Applicable" = "gray"),
    na.translate = FALSE) +
  ggspatial::annotation_north_arrow(location = "br") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    legend.box = "horizontal")
```
